\ProjectName{}: Data Imputation as a Query Optimization

### 2017-02-18 paper/algorithm.tex
paper: added todos
paper: removed histogram helpers and tried to rework into simpler form
paper: fixed general typos, added clearer todos

The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
Figure~\ref{fig:query-schematic} shows the layout of a query involving three tables, absent any imputation operators.

@@

For each sub-query, we keep track of the estimated number of
null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
the sub-query. We use this estimate to compute
$\text{Loss}(Q)$ as follows.

\begin{align*}
  \textsc{Loss}(q) = \begin{cases}
    \sum_{c \in C} \textsc{Missing}(c, q') & q = \delta_C(q') \\
    \frac{1}{\sqrt{|q'|}} \sum_{c \in C} \textsc{Missing}(c, q') & q = \mu_C(q') \\
    \textsc{Loss}(q_1') + \textsc{Loss}(q_2') & q = q_1' \bowtie_\psi q_2' \\
    \textsc{Loss}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $p \in (0, 1]$ which decreases as the number of
tuples available to train the imputation algorithm increases.  Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information loss \textit{decreases} with more complete values to work
with. The inverse square loss can be seen as a heuristic first approximation to the learning guarantees of
regression trees. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.

@@

Evaluation of the time complexity for \textit{Impute} is thornier, as the properties of
different decision tree building algorithms can vary significantly, the underlying IO cost
is not easily extracted (especially in the case that the entire stream of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans), though the range of $\alpha$ that does promote tradeoffs is condensed.

@@

\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned}
                                           Q \cup \{(T,\ \argmin_{q \in P}\ \textsc{Cost}(q))\}\ \text{where} \\
                                           P = \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p) 
                                                                   \end{aligned}
                                           \right\}
                                         \end{aligned}
    % \in 
  \end{align*}
  \caption{Semantics of the plan cache.}
\end{figure}

The query planner (Algorithm~\ref{algo:top-level-planner}), operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan.
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache

@@

Algorithm~\ref{algo:plan-helpers} presents a series of helper functions used by the top level planner, shown in Algorithm~\ref{algo:top-level-planner}.

\begin{algorithm}
\input{algorithms/plan-helpers.tex}
\end{algorithm}

@@

\begin{algorithm}
\input{algorithms/histogram-helpers.tex}
\end{algorithm}

@@

\subsection{Complexity}
\todo{this section needs to be cleaned up/clarified}
Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
note that if we remove our restriction on types of imputation($\delta_{min}, \mu_{min}, \mu_{max}$), and allow any arbitrary subset of attributes to be imputed,
then every single operator in the query plan has a number of imputations exponential in the number of dirty columns. Our restriction, instead, increases the number
of plans (in the worst case) at each operator by a factor of 3. Of course, this implies that in the worst case (where all dirty sets tracked are distinct throughout the query plan),
we explore a number of plans that further scales the number of plans in the original algorithm by an exponential factor.

>>>>

The space of query plans is similar to that considered by System R~\cite{blasgen1981system}, with the addition of imputation operators appearing before and after traditional operators.
Figure~\ref{fig:query-schematic} shows a plan schematic for a query involving three tables, absent any imputation operators.

@@

\todobox{this section must be rewritten to show new approach}{
For each sub-query, we keep track of the estimated number of
null values in a column  ($\textsc{Missing}(c, Q)$) by using the histograms associated with
the sub-query. We use this estimate to compute
$\text{Loss}(Q)$ as follows.

\begin{align*}
  \textsc{Loss}(q) = \begin{cases}
    \sum_{c \in C} \textsc{Missing}(c, q') & q = \delta_C(q') \\
    \frac{1}{\sqrt{|q'|}} \sum_{c \in C} \textsc{Missing}(c, q') & q = \mu_C(q') \\
    \textsc{Loss}(q_1') + \textsc{Loss}(q_2') & q = q_1' \bowtie_\psi q_2' \\
    \textsc{Loss}(q') & q = \sigma_\phi(q'), \\ & q = \pi_C(q') \\
  \end{cases}
\end{align*}
Dropping tuples which contain a null value incurs a loss of 1 for each null field dropped.
Imputing null fields incurs a loss penalty $p \in (0, 1]$ which decreases as the number of
tuples available to train the imputation algorithm increases.  Intuitively, imputation
accuracy \textit{increases} when more complete attributes and complete tuples are available;
correspondingly, the information loss \textit{decreases} with more complete values to work
with. The inverse square loss can be seen as a heuristic first approximation to the learning guarantees of
regression trees. Since we conceive that
alternate imputation strategies could be used, the analyst may provide a corresponding
loss with a reasonable parameterization.
}

@@

\todobox{This seems tailored to decision trees, but we should talk more generally since
we claim the imputation step isn't the focus}{
Evaluation of the time complexity for \textit{Impute} is thornier, as the properties of
different decision tree building algorithms can vary significantly, the underlying IO cost
is not easily extracted (especially in the case that the entire stream of tuples does not
fit in memory), and the CPU cost dominates (in contrast to the other operators, which may
not even consider CPU cost explicitly). For example, \cite{martin1995time} find that the
time complexity of the \textit{build tree} algorithm for one commonly-used class of
decision trees is a function of the number of classes, several overhead constants, the
parameterization of the partition and heuristic functions, and the \textit{arity} (the
number of subsets considered for each split). Indeed, the \textit{build tree} phase often
does not even dominate computation, as post-processing steps like pruning, which are vital
in achieving good performance, can have cost exponential in the height of the tree. In
practice, we find that a simple parameterization can be acceptable (i.e. yield intelligent query
plans), though the range of $\alpha$ that does promote tradeoffs is condensed.
}

@@

\todo{formatting of plan cache semantics needs to be fixed}
\begin{figure}
  \begin{align*}
    \llbracket Q[T] \rrbracket &= \begin{cases}
      P & (T, P) \in Q \\
      \emptyset & \text{otherwise} \\
    \end{cases} \\
    \llbracket Q[T] \lhd p \rrbracket &= \begin{aligned}
                                           Q \cup \{(T,\ \argmin_{q \in P}\ \textsc{Cost}(q))\}\ \text{where} \\
                                           P = \left\{q ~|~ \begin{aligned}
                                                                     q \in \llbracket Q[T] \rrbracket \cup \{p\}, \\
                                                                     \textsc{Dirty}(q) = \textsc{Dirty}(p) 
                                                                   \end{aligned}
                                           \right\}
                                         \end{aligned}
    % \in 
  \end{align*}
  \caption{Semantics of the plan cache.}
\end{figure}

\todobox{reorder these and explain algorithm, which right now is cutoff}{
The query planner (Algorithm~\ref{algo:top-level-planner}), operates as follows.
First, it collects the set of attributes which are used by the operators in the plan or which are visible in the output of the plan.
This set will be used to determine which attributes are imputed.
Then, it constructs a plan cache

@@

Algorithm~\ref{algo:plan-helpers} presents a series of helper functions used by the top level planner, shown in Algorithm~\ref{algo:top-level-planner}.
}
\begin{algorithm}
\todo{fix input descriptions and small error in optjoin}
\input{algorithms/plan-helpers.tex}
\end{algorithm}

@@

%\begin{algorithm}
%\input{algorithms/histogram-helpers.tex}
%\end{algorithm}

@@

\subsection{Complexity}
\todo{this section needs to be cleaned up/clarified}
Our optimization algorithm builds off the approach taken by System R \cite{blasgen1981system}, therefore our algorithm still operates in exponential time. Indeed, 
note that if we remove our restriction on types of imputation (i.e. minimal vs maximal), and allow any arbitrary subset of attributes to be imputed,
then every single operator in the query plan has a number of imputations exponential in the number of dirty columns. Our restriction, instead, increases the number
of plans (in the worst case) at each operator by a factor of 3. Of course, this implies that in the worst case (where all dirty sets tracked are distinct throughout the query plan),
we explore a number of plans that further scales the number of plans in the original algorithm by an exponential factor.

### 2017-02-18 paper/conclusion.tex
paper: fixed general typos, added clearer todos

\section{Conclusion}
As shown in \ProjectName{}, missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
information loss and runtime. We showed that different values for the tradeoff parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with \textit{CDC} and \textit{freeCodeCamp} survey-based datasets
showed that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query result in execution that
takes a fraction of the time compared to the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be minimal. Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

@@

\textbf{Cost model intepretation:} While the current system
provides a parameter ($\alpha$) to tradeoff between imputation quality and runtime,
there is no direct interpretation of its unit-less value. One natural extension to this work
is to reformulate the cost-model in an interpretable fashion.

>>>>

\section{Conclusion}
As shown in \ProjectName{}, missing values and their imputation can be successfully integrated into the relational calculus and
existing cost-based plan optimization frameworks. We implement imputation actions, such as dropping or imputing values with a machine
learning technique, as operators in the algebra and use a simple, but effective, cost model to consider tradeoffs in 
information loss and runtime. We showed that different values for the tradeoff parameter can yield substantially
different plans, allowing the user to express their preferences for performance on a per-query basis.
Our experiments with \textit{CDC} and \textit{freeCodeCamp} survey-based datasets
showed that \ProjectName{} can be used successfully with real-world data, expressing standard SQL queries over data with missing values. 
We craft a series of realistic queries, meant to resemble the kind of questions an analyst might ask during
the data exploration phase. The plans selected for each query execute in a fraction of the time compared to the standard approach of imputing on the entire base tables and
then formulating queries. Furthermore, the difference in query results between the two approaches is
shown to be minimal. Data analysts need not commit to a specific imputation strategy and can instead
vary this across queries.

@@

\textbf{Cost model interpretation:} While the current system
provides a parameter ($\alpha$) to tradeoff between imputation quality and runtime,
there is no direct interpretation of its unit-less value. One natural extension to this work
is to reformulate the cost-model in an interpretable fashion.

### 2017-02-18 paper/examples.tex
paper: removed search space size todo note in examples
paper: fix mistake in figure
paper: fixed general typos, added clearer todos

While careful modeling techniques will be necessary before putting
results into the final research paper, the epidemiologist is anxious
to get a quick and accurate view into the data. However,
the data has been collected through CDC surveys (see Section~\ref{subsec:datasets} for details),
and there is a significant amount of missing data across all
relevant attributes. The researcher must develop a strategy
for handling the missing values. She could drop any records with
missing values and continue her analysis with the remaining data,
or she could fit a model on the base tables which can take advantage of existing correlations between attributes to fill in the missing values.
If she drops the records with missing data, she will throw away the majority of her data set, as many records have missing fields.
Fitting a model on the dataset will take more programming effort and may take a long time to run.

@@

Rather than work through
the implications of each case, she submits a query, as shown in Figure~\ref{fig:example-query}, to ImputeDB.\@ The system, parameterized by a value indicating
a tradeoff between information loss and runtime cost, then finds
the optimal query plan. Figure~\ref{fig:quality-plan} and Figure~\ref{fig:fast-plan} show what this plan would be
when optimizing for quality and performance, respectively. These two plans are found within a broader search space
of XXXX plans. \todo{add a small comment about the number of plans in the plancache
to highlight size of search space}.

The differences between the two plans highlight the subtleties of optimizing queries which contain imputation operators.
Note that imputation operators are placed in different points of the plan, operating over different columns, and indeed
even influencing the order of joins. The quality-optimized plan (Figure~\ref{fig:quality-plan}) uses the impute operation, rather
than dropping tuples with missing values. It waits to impute \verb|demo.income| until after the final join has taken place, but
other imputations take place earlier on in the plan, some before filtering and join operations. 
Meanwhile, the performance-optimized plan (Figure~\ref{fig:fast-plan}) performs a single impute operation (\verb|exams.weight|), and applies drops to tuples with missing values before aggregating.

@@

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$
        [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
          [.$\bowtie_{\text{demo.id} = \text{exams.id}}$
            [.$\sigma_{\text{demo.gender} = 2}$ demo ]
            [.$\sigma_{\text{exams.weight} \geq 120}$ [.$\mu_{\text{exams.weight}}$ exams ] ] ] labs ] ] ] ]
\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for performance.}
\label{fig:fast-plan}
\end{figure}

Tens of queries later, the epidemiologist has the holistic view of the
data that they required before carefully crafting their own tailored
imputation model. Knowing that there may be a need to explore
this data set further, they can easily incorporate their imputation model
into ImputeDB for future use.

>>>>

While careful modeling techniques will be necessary before putting
results into the final research paper, the epidemiologist is anxious
to get a quick and accurate view into the data. However,
the data has been collected through CDC surveys (see Sec.~\ref{subsec:datasets} for details),
and there is a significant amount of missing data across all
relevant attributes. The researcher must develop a strategy
for handling the missing values. She could drop any records with
missing values and continue her analysis with the remaining data,
or she could fit a model on the base tables which can take advantage of existing correlations between attributes to fill in the missing values.
If she drops the records with missing data, she will throw away the majority of her data set, as many records have missing fields.
Fitting a model on the dataset will take more programming effort and may take a long time to run.

@@

Rather than work through
the implications of each case, she submits a query, as shown in Figure~\ref{fig:example-query}, to ImputeDB.\@ The system, parameterized by a value indicating
a tradeoff between information loss and runtime cost, then finds
the optimal query plan. Figures~\ref{fig:quality-plan} and ~\ref{fig:fast-plan} show what this plan would be
when optimizing for quality and performance, respectively.

The differences between the two plans highlight the subtleties of optimizing queries which contain imputation operators.
Note that imputation operators are placed in different points of the plan, operating over different columns, and 
even influencing the order of joins. The quality-optimized plan (Figure~\ref{fig:quality-plan}) uses the impute operation, rather
than dropping tuples with missing values. It waits to impute \verb|demo.income| until after the final join has taken place, but
other imputations take place earlier on in the plan, some before filtering and join operations. 
Meanwhile, the performance-optimized plan (Figure~\ref{fig:fast-plan}) performs a single impute operation (\verb|exams.weight|), and applies drops to tuples with missing values before aggregating.

@@

\begin{figure}
  \Tree
  [.$\pi_{\text{income, AVG(white\_blood\_cell\_ct)}}$
    [.$g_{\text{income, AVG(white\_blood\_cell\_ct)}}$
      [.$\delta_{\text{demo.income, labs.white\_blood\_cell\_ct}}$
        [.$\bowtie_{\text{exams.id} = \text{labs.id}}$
          [.$\bowtie_{\text{demo.id} = \text{exams.id}}$
            [.$\sigma_{\text{demo.gender} = 2}$ demo ]
            [.$\sigma_{\text{exams.weight} \geq 120}$ [.$\delta_{\text{exams.weight}}$ exams ] ] ] labs ] ] ] ]
\caption{A plan for the query in Fig.~\ref{fig:example-query}, optimized for performance.}
\label{fig:fast-plan}
\end{figure}

Tens of queries later, the epidemiologist has the holistic view of the
data that she required before carefully crafting her own tailored
imputation model. Knowing that there may be a need to explore
this data set further, she can easily incorporate their imputation model
into ImputeDB for future use.

### 2017-02-18 paper/experiments.tex
paper: fixed general typos, added clearer todos

\section{Experiments}
To evaluate the performance of our system, we generate and execute plans for queries
for two separate survey-based data sets, showing that our system
is ideal for early dataset exploration.

@@

The queries on the CDC NHANES data consists not only of projections and selections,
but also interesting joins and aggregations. Our aim was to craft meaningful queries
that would provide performance figures relevant to practitioners who may use
similar datasets.

Query~\ref{q1} calculates the average weight for users based on their income data.
To do so, it joins demographics and physical examination data. Query~\ref{q2}
compares cholesterol levels for individuals who are both ends of the income
spectrum and who satisfy a minimum weight requirement. Query~\ref{q3}
extracts the maximum blood lead levels for subjects who satisfy a particular
waist circumference requirement.

In Query~\ref{q4}, we calculate the average income for survey participants,
based on their bootcamp attendance. Query~\ref{q5} estimates the average
age of women from the Uniter States who participated. Query~\ref{q6} 
calculates the average amount of money survey participants with student debt destined to
learning based on their school degree. Finally, in Query~\ref{q7}, we join the FCC data with a reference table provided
by the World Bank which summarizes GDP per Capita across various
countries\cite{worldbank-data}. The query calculates the average GDP-per-capita for participants based on whether or not they
attended a bootcamp.

@@

\begin{table*}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on FCC data}
 \label{fig:queries-fcc}
 \end{subtable}
  \caption{Queries used in our experiments.}
  \label{fig:queries}
\end{table*}

@@

\todo{actually run experiments and rewrite this section}
Table~\ref{fig:experiments} provides a summary of our experimental results.
We run each query in four configurations: without imputation on clean data, without imputation on dirty data, and with imputation on dirty data for $\alpha = 0$ and $\alpha = 1$.
When a query returns a relation of the same ``shape'' (same number of tuples and same schema) on both clean and dirty data, we can compute an error value.
For our experiments, we report root-mean-square error (RMSE) for the base error (i.e. the error in the dirty data without imputation with respect to the clean data) and the change
in RMSE for the imputed plans.
For imputations, we run queries using the two extreme values of $\alpha$: 0 and 1, representing a focus on runtime and on information loss, respectively.
For each query, the table displays the base error, the change in error in both imputation scenarios, and the running time in seconds.

@@

Finally, we performed CE-CART imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
We were unable to complete the full imputation due to resource constraints, but the process ran for more than one hour without terminating.
This time compares favorably to our maximum of 72 seconds for a query with imputation on the ACS data.
Given the varying needs of database users, \ProjectName{} clearly allows flexibility without a significant performance hit.

>>>>

\section{Experiments}
To evaluate the performance of our approach,  we implemented
a prototype system in Java, following a traditional iterator model.
For our experiments we plan and execute queries
for two separate survey-based data sets, showing that our system
is ideal for early dataset exploration.

@@

The queries on the CDC NHANES data consist not only of projections and selections,
but also interesting joins and aggregations. Our aim was to craft meaningful queries
that would provide performance figures relevant to practitioners who may use
similar datasets.

Queries 1-3 are on the CDC data. Query~\ref{q1} calculates the average weight for users based on their income data.
To do so, it joins demographics and physical examination data. Query~\ref{q2}
compares cholesterol levels for individuals who are both ends of the income
spectrum and who satisfy a minimum weight requirement. Query~\ref{q3}
extracts the maximum blood lead levels for subjects who satisfy a particular
waist circumference requirement.

Queries 4-7 are on the FCC data. In Query~\ref{q4}, we calculate the average income for survey participants,
based on their bootcamp attendance. Query~\ref{q5} estimates the average
age of women from the Uniter States who participated. Query~\ref{q6} 
calculates the average amount of money survey participants with student debt destined to
learning based on their school degree. Finally, in Query~\ref{q7}, we join the FCC data with a reference table provided
by the World Bank which summarizes GDP per Capita across various
countries\cite{worldbank-data}. The query calculates the average GDP-per-capita for participants based on whether or not they
attended a bootcamp.

@@

\begin{table*}
\todo{clean up formatting here}
\centering
 \begin{subtable}{\linewidth}
  \input{queries-cdc}
  \caption{Queries on CDC data}
  \label{fig:queries-cdc}
 \end{subtable}
 ~
 \begin{subtable}{\linewidth}
 \input{queries-fcc}
 \caption{Queries on FCC data}
 \label{fig:queries-fcc}
 \end{subtable}
  \caption{Queries used in our experiments.}
  \label{fig:queries}
\end{table*}

@@

\todobox{actually run experiments and rewrite this section}{
Table~\ref{fig:experiments} provides a summary of our experimental results.
We run each query in four configurations: without imputation on clean data, without imputation on dirty data, and with imputation on dirty data for $\alpha = 0$ and $\alpha = 1$.
When a query returns a relation of the same ``shape'' (same number of tuples and same schema) on both clean and dirty data, we can compute an error value.
For our experiments, we report root-mean-square error (RMSE) for the base error (i.e. the error in the dirty data without imputation with respect to the clean data) and the change
in RMSE for the imputed plans.
For imputations, we run queries using the two extreme values of $\alpha$: 0 and 1, representing a focus on runtime and on information loss, respectively.
For each query, the table displays the base error, the change in error in both imputation scenarios, and the running time in seconds.

@@

Finally, we performed CE-CART imputation on the entirety of the ACS table, in order to compare our system to the traditional, pre-processing approach to imputation.
We were unable to complete the full imputation due to resource constraints, but the process ran for more than one hour without terminating.
This time compares favorably to our maximum of 72 seconds for a query with imputation on the ACS data.
Given the varying needs of database users, \ProjectName{} clearly allows flexibility without a significant performance hit.
}

\todo{Add timing comparison to ACS table since that should take longer than base on the data we have}

### 2017-02-18 paper/introduction.tex
paper: fixed general typos, added clearer todos

Handling incorrect or dirty data is a complex and challenging problem for analysts.
According to some estimates, data scientists spend between 50 and 80 percent of their
time performing tasks such as collecting and cleaning datasets\cite{data-science-cleaning}.
One way in which a dataset can be dirty is for parts of it to be missing altogether.
Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
To handle this problem, users may manually clean their dataset by performing
some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
Although manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
Additionally, it may not even be necessary to completely clean the data to make it usable.
Some users may be willing to run queries on dirty data, simply ignoring the missing values, as long as they do not have to pay the cost of imputation.
Others may want to run queries on a subset of the data, and so do not need to impute
every field in every record. Yet others may want to customize the 
imputation algorithm for the tradeoffs and demands of a particular domain.

In this paper, we present \ProjectName{}, a database system which is designed to interact with a dirty dataset as though it were clean.
The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
To achieve this goal, we perform imputation on the fly, during query execution.
Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows users to flexibly trade accuracy for computation time.

>>>>

Handling incorrect or dirty data is a complex and challenging problem for analysts.
According to some estimates, data scientists spend between 50 and 80 percent of their
time performing tasks such as collecting and cleaning datasets\cite{data-science-cleaning}.
One way in which a dataset can be dirty is for parts of it to be missing altogether.
Missing data is one of the simpler variants of dirty data, but if handled naively, can still cause analyses to be incorrect.
To handle this problem, users may manually clean their dataset by performing
some statistical analysis to replace missing data elements with likely values.
This process is called \emph{imputation}.
Traditional imputation methods replace all missing values in a dataset to create a new clean dataset.
Although this manual imputation solves the problem of missing data, in the age of big data it may be very expensive to run an imputation algorithm on an entire dataset.
Additionally, it may not even be necessary to completely clean the data to make it usable.
Some users may be willing to run queries on dirty data, simply ignoring any missing values, as long as they do not have to pay the cost of imputation.
Others may want to run queries on a subset of the data, and so do not need to impute
every field in every record. Yet others may want to customize the 
imputation algorithm for the tradeoffs and demands of a particular domain.

In this paper, we present \ProjectName{}, a database system which is designed to interact with a dirty dataset as though it were clean.
The guiding design principle behind \ProjectName{} is that the user should never see missing data or have to modify their queries to account for it.
To achieve this goal, we perform imputation on the fly, during query execution.
Performing imputation at query time allows our system to impute only the data necessary to run the query, and it allows users to flexibly trade imputation quality for computation time.

### 2017-02-18 paper/relatedwork.tex
paper: fixed general typos, added clearer todos

\subsection{Missing Values and Databases}
There is a long history in the database community surrounding the
treatment of nulls. As early as 1973, \cite{codd1973understanding}
provides a treatment of the semantics of null. Multiple
papers have described various (at times conflicting) treatments
of nulls \cite{grant1977null}. \ProjectName's main design invariant - that no relational operator
see attributes with missing values for attributes it must operate on and users should never see
missing data - eliminates
the need to handle null value semantics, while guaranteeing soundness (modulo
imputation strategy) of the query evaluation.

>>>>

\subsection{Missing Values and Databases}
There is a long history in the database community surrounding the
treatment of nulls. As early as 1973, \cite{codd1973understanding}
provides a treatment of the semantics of null. Multiple
papers have described various (at times conflicting) treatments
of nulls \cite{grant1977null}. \ProjectName's main design invariant - no relational operator
sees missing values for attributes it must operate on and users should never see
missing data - eliminates
the need to handle null value semantics, while guaranteeing soundness (modulo
imputation strategy) of the query evaluation.
